---
title: "Personal Activity Analysis"
author: "Chris Sirico"
date: "11/27/2017"
output: 
  github_document:
    html_preview: true
---
<!-- ☝️Trying above YAML for updated github-friendly knitr output; code_folding
might not work; might have to use:
    smart: false
    if pandoc causes problems -->
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, fig.retina=4, warning=FALSE, message=FALSE)
activity <- readr::read_csv("activity.csv") # read in data as tibble
library(dplyr)
library(ggplot2)

# "steps" is number of steps measured in that interval
# "data" is the date in form YYYY-MM-DD
# "interval" has minutes in the rightmost digits and hours in leftmost digits
# let's at a day-of-week column:
activity <- mutate(activity,
              day = weekdays(date))

# activity$date %>% max() - activity$date %>% min()
#> Time difference of 60 days (length of the study)

# ASSIGNMENT

# - Code for reading in the dataset and/or processing the data
# - Histogram of the total number of steps taken each day
# - Mean and median number of steps taken each day
# - Time series plot of the average number of steps taken
    # continuous, start-to-end
    # by weekday, cumulative
# - The 5-minute interval that, on average, contains the maximum number of steps
    # by weekday
    # by day
# • Code to describe and show a strategy for imputing missing data
    # remove rows w/ missing data for some steps; where is this problematic?
    # take average by weekday and interval of day to impute for others
# • Histogram of the total number of steps taken each day after missing values
# are imputed; kind of already done
# • Panel plot comparing the average number of steps taken per 5-minute interval
# across weekdays and weekends
# - All of the R code needed to reproduce the results (numbers, plots, etc.) in
# the report

# function to emulate ggplot hues
#https://stackoverflow.com/questions/8197559/emulate-ggplot2-default-color-palette
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}

```

`r scales::percent(mean(is.na(activity$steps)))` of steps data is missing. We may have to impute those data or discard them.

```{r summary}
activity$steps %>% summary()

# removing 0 step rows
activity %>% 
  filter(steps != 0) %>%
  select(steps) %>%
  summary()
```
It's interesting to note that the maximum number of steps in a 5-minute interval is 806. That's 2.68 steps per second, on par with a tempo of [161 beats per minute](https://www.youtube.com/watch?v=E2X4Fv4FjE4). That would seem to be faster than a steady jog, perhaps just short of a sprint—high-intensity exercise, regardless of the activity.

```{r}
arrange(activity, desc(steps))
```
Looking at the highest number of steps per interval, it seems like our subject works out early in the morning.

There are 8 dates for which data is missing.
```{r}
no_step <- activity %>%
  group_by(date) %>%
  summarise(steps = sum(steps, na.rm = F)) %>%
  filter(is.na(steps)) %>%
  arrange(date)

activity_na_rm <- activity %>%
  filter(date != no_step$date)
```


```{r prop, fig.width=6, fig.height=2}
# show proportion of movement to sedentary intervals
# Stacked barplot -- pretty hacky way of getting there

prop <- activity_na_rm %>%
  filter(!is.na(steps))  # filter out missing values

total <- nrow(prop)

prop <- prop %>%
  mutate(
    movement = steps > 0,
    movement2 = "active"
            ) %>%
  group_by(movement, movement2) %>%
  summarize(active = n(),
            prop = active / total
            )

# Calculate the cumulative sum of steps
prop <- plyr::ddply(prop, "movement2",
                   transform, label_pos=cumsum(active))

# Set label locations
prop$label_pos[1] <- 8900
prop$label_pos[2] <- 1450

ggplot(prop, aes(x=movement2, y=active, fill=movement)) +
  geom_bar(stat = "identity") +
  #           color="white", size=3.5)+
  # scale_fill_brewer(palette="Paired")+
  coord_flip() +
  labs(title = "Subject was active in 28% of intervals",
    x = NULL,
    y = "Intervals"
  ) +
# active label
  geom_text(data = tibble::as.tibble(split(prop, prop$movement)[[1]]), # use only one row of data
          aes(y=label_pos[1],
              x=movement2[1],
              label=stringr::str_c(as.character(scales::percent(prop[1])), "\nsedentary")
          ),
          hjust = 0) +
  # sedentary label
  geom_text(data = tibble::as.tibble(split(prop, prop$movement)[[2]]), # use only one row of data
            aes(y=label_pos[1],
                x=movement2[1],
                label=stringr::str_c(as.character(scales::percent(prop[1])), "\nactive")
            ),
            hjust = 0) +
  # theme_classic() +
  # nix ticks & hide legend
  theme(axis.text.y = element_blank(),
  axis.ticks.y = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  panel.background = element_blank(),
  plot.title = element_text(hjust = .09)) +
  # clean theme
  theme(legend.position="none")
```

```{r histo}
byday <- activity_na_rm %>%
  group_by(date) %>%
  summarise(steps = sum(steps, na.rm = TRUE)) # filtering out missing values

ggplot(byday) +
geom_histogram(aes(steps), binwidth = 600) +
# vertical x labels
theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
labs(title = "Subject achieved about 10,000 steps per day",
  x = "number of steps/day",
  y = "number of occurences"
) +  # dashed median and mean lines
geom_vline(aes(
  xintercept = median(byday$steps, na.rm = T),
  color = "Median"),
    linetype = "dashed"
) +
geom_vline(aes(
  xintercept = mean(byday$steps, na.rm = T),
  color = "Mean"),
    linetype = "dashed"
) +
scale_y_continuous(breaks = c(1:10),
                   minor_breaks = NULL) +
 geom_text(data = tibble::tibble(steps = c(0, max(byday$steps)),
                date = c(1,2)),
             aes(y=9,
               x=13000,
               label=stringr::str_c("Median:\n", as.character(median(byday$steps, na.rm = T)))
            ),
           color = "#129496"
           # ,
           #  hjust = 0
           ) +
geom_text(data = tibble::tibble(steps = c(0, max(byday$steps)),
                date = c(1,2)),
             aes(y=5,
               x=6000,
               label=stringr::str_c("Mean:\n", as.character(round(mean(byday$steps, na.rm = T), 2)))
            ),
           color = "#E7655C"
           # ,
           #  hjust = 0
           ) +
  theme(legend.position="none")


```

The study's subject achieves the recommended [10,000 steps](https://www.ncbi.nlm.nih.gov/pubmed/11131268) on most days.

```{r long-time-series}
steps_seq <- activity
steps_seq <- tibble::rowid_to_column(steps_seq, "ID")

# I want to set up x axis tick labels with the number of days instead of intervals
# 288 5-minute intervals = 1 day
# Not sure how to get this to work

# intervals <- as.character(seq(from = 0, to = max(steps_seq$ID), by = 288))
# days <- as.character(seq(from = 0, to = 61, by = 1))
# day_labels <- vector("character", 62)
# day_labels <- for(i in seq_along(day_labels)) {
#   day_labels[i] <- stringr::str_c("'", intervals[i], "'", " = ", "'", days[i], "'")
# }

#  output should look like:
  # c("0" = "1", "288" = "2", "576" = "3", ...)

ggplot(data = steps_seq, aes(ID, steps)) + 
  geom_line() +
  labs(title = "Time series over 60-day study",
       subtitle = "Data were missing on 8 days",
  x = "number of steps/day",
  y = "number of occurences"
) +
  scale_x_discrete(breaks = c(seq(from = 0, to = max(steps_seq$ID), by = 288))
      # labels = day_labels  # still haven't figured out how to customize labels
                   ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5))
```

```{r week-time-series, include= FALSE}
# not including this for now -- goal is to show average interval step values grouped by week
# hours are broken up into because they're treated like hundreds in interval column
# plot is showing 1 day right now
# not sure how to format data to get x axis showing days-of-week and hours
steps_week <- steps_seq
steps_week$datetime <- with(steps_week, lubridate::ymd_hm(stringr::str_c(date, interval)))
steps_week$week <- steps_week$ID %/% 2016  # create week number column
steps_week %>%
  group_by(week, interval) %>%
  summarise(steps = mean(steps, na.rm=TRUE)) %>%  # removing missing values here
# unite(week_int, week, interval, sep = "-") %>%
  
ggplot(aes(interval, steps)) + 
  geom_line() +
  labs(title = "Time series over 60-day study",
       subtitle = "Data were missing on 8 days",
  x = "number of steps/day",
  y = "number of occurences"
) +
  # scale_x_discrete(breaks = "")
  #     # labels = day_labels  # still haven't figured out how to customize labels
  #                  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5))

```

```{r max-interval}
daily_avg <- activity %>%
  group_by(interval) %>%
  summarise(steps = mean(steps, na.rm = TRUE)) %>%
  arrange(desc(steps))

ggplot(daily_avg, aes(interval, steps)) +
  geom_area() + 
  geom_point(aes(835, 206), color = "red") +
  geom_text(data = daily_avg[1,1], aes(950, 200), label = "max interval: 8:35 am\naverage steps: 206", hjust = 0) +
  labs(title = "Daily average steps per interval - peak @ 8:35 am"
    
  )
# Might be fun to come back and do this by interval and weekday
```



```{r histo-f-ckup, include=FALSE}
# excluding this chunk since it doesn't track days, according to the prompt

# zero-less

# zeroless <- activity %>% 
# filter(steps != 0)
#   
# ggplot(zeroless) +
# geom_histogram(aes(steps),
#                bins = 100) +
# scale_x_continuous(breaks = seq(0, max(activity$steps, na.rm = T), by = 25)) +
# theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
# labs(title = "Occurence of numbers of steps",
#   subtitle = "(Zeros removed)",
#   x = "number of steps",
#   y = "number of intervals"
# ) +
# geom_vline(aes(
#   xintercept = median(zeroless$steps, na.rm = T),
#   color = "Median"),
#     linetype = "dashed"
# ) +
# geom_vline(aes(
#   xintercept = mean(zeroless$steps, na.rm = T),
#   color = "Mean"),
#     linetype = "dashed"
# )
# 
# # zero-less square root scale
# ggplot(zeroless) +
# geom_histogram(aes(steps),
#                bins = 100) +
# scale_x_sqrt(
#   breaks = c(
#   seq(0, 25, by = 5),
#   seq(50, 400, by = 25),
#   seq(450, max(activity$steps, na.rm = T), by = 50)
#   ),
#   minor_breaks = c(
#   seq(0, 45, by = 5),
#   seq(50, 195, by = 5),
#   seq(200, 480, by = 25),
#   seq(450, max(activity$steps, na.rm = T), by = 25)
#   )
# ) +
# theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
# labs(title = "Occurence of numbers of steps",
#   subtitle = "(Zeros removed, square root scale)",
#   x = "number of steps",
#   y = "number of intervals"
# ) +
#   geom_vline(aes(
#   xintercept = median(zeroless$steps, na.rm = T),
#   color = "Median"),
#     linetype = "dashed"
# ) +
# geom_vline(aes(
#   xintercept = mean(zeroless$steps, na.rm = T),
#   color = "Mean"),
#     linetype = "dashed"
# )
```