---
title: "Personal Activity Analysis"
author: "Chris Sirico"
date: "11/27/2017"
output: 
  github_document:
    html_preview: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, fig.retina=4, warning=FALSE, message=FALSE)
activity <- readr::read_csv("activity.csv") # read in data as tibble
library(dplyr)
library(ggplot2)

# "steps" is number of steps measured in that interval
# "data" is the date in form YYYY-MM-DD
# "interval" has minutes in the rightmost digits and hours in leftmost digits
# let's create a day-of-week, hour and minute column:
activity <- mutate(activity,
              day = weekdays(date),
              hour = interval %/% 100,
              minute = interval %% 100)

# activity$date %>% max() - activity$date %>% min()
#> Time difference of 60 days (length of the study)

# ASSIGNMENT

# - Code for reading in the dataset and/or processing the data
# - Histogram of the total number of steps taken each day
# - Mean and median number of steps taken each day
# - Time series plot of the average number of steps taken
    # continuous, start-to-end
    # by weekday, cumulative
# - The 5-minute interval that, on average, contains the maximum number of steps
    # by weekday
    # by day
# • Code to describe and show a strategy for imputing missing data
    # remove rows w/ missing data for some steps; where is this problematic?
    # take average by weekday and interval of day to impute for others
# • Histogram of the total number of steps taken each day after missing values
# are imputed; kind of already done
# • Panel plot comparing the average number of steps taken per 5-minute interval
# across weekdays and weekends
# - All of the R code needed to reproduce the results (numbers, plots, etc.) in
# the report
```

This is an exploration of pedometer data collected on one subject over a 2-month study.

Let's preview our data.

```{r preview, echo = "FALSE"}
print(activity)
```

The variables are `date`, 5-minute `interval` (as a time of day), and `steps` taken per interval. A day-of-week column has also been added (`day`), and interval has been separated into additional `hour` and `minute` columns.

`r scales::percent(mean(is.na(activity$steps)))` of `steps` data are missing, or `r sum(is.na(activity$steps))` values. A look at the data reveals 8 dates for which those data are completely missing, but there are no missing values outside those dates.

Those dates can simply be removed from the dataset for most of this exploration. This will still render accurate means, medians and average `step` counts per day. 

Any analysis that involves a count and day-of-week (e.g. number of steps taken on Tuesdays), however, would require imputation. A possible option for imputation in such case would be the median `step` value taken by the corresponding interval (time of day) and day-of-week. See the `impute` chunk for code to carry out the imputation.

```{r na-rm}
# find days with missing data
no_step <- activity %>%
  group_by(date, day) %>%
  summarise(steps = sum(steps, na.rm = FALSE)) %>%
  filter(is.na(steps)) %>%
  arrange(date)

# remove dates with missing values from dataset
activity_na_rm <- activity %>%
  filter(!(date %in% no_step$date))

# plot missing days
activity %>%
  group_by(date) %>%
  summarize(nas = sum(is.na(steps))) %>%
ggplot(aes(date, nas))+
  geom_col() +
  scale_x_date(date_breaks = "1 day") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
  labs(title = "Missing values by day",
       subtitle="8 days completely missing; no partially missing days.",
       y = "Number of 5-minute intervals missing")
```

```{r impute}
# IMPUTE
# impute median values for missing days

# puts missing values at bottom of dataframe, in date order
imputed <- activity %>%
  arrange(steps)

# summarize by median
activity_medians <- activity_na_rm %>%
  group_by(day, interval) %>%
  summarize(steps = median(steps)) %>%
  tidyr::spread(day, steps)

# number of missing values
missing_total <- sum(is.na(activity$steps))

# getting an error when I subset:
#> 1: Unknown or uninitialised column: 'steps'.
# trying this to fix that
# col_names <- names(imputed)
# names(imputed) <- col_names

# find first missing value
na_start<-(nrow(imputed)-missing_total+1)

# start with first missing value (row 15265)
# get day of week
# find day in activity_medians
# replace steps with median_steps day column for 1 day's rows (288)
# go to next row and get day
# iterate to end of dataframe
for (i in seq(na_start, nrow(imputed), by = 288)){
  for (j in seq(0, 287)){
    imputed$steps[i+j] <- activity_medians[[(j+1),(imputed$day[(i+j)])]]
    
  }
}
```

Below we see summary statistics for our data with and without zero values removed.
```{r summary}
activity_na_rm$steps %>% summary()

# removing 0 step rows
activity_na_rm %>% 
  filter(steps != 0) %>%
  select(steps) %>%
  summary()
```
It's interesting to note that the maximum number of steps in a 5-minute interval is 806. That's 2.68 steps per second, on par with a tempo of [161 beats per minute](https://www.youtube.com/watch?v=E2X4Fv4FjE4). That would seem to be faster than a steady jog and perhaps just short of a sprint. That would seem to be quite high-intensity exercise, regardless of the specific activity.

```{r max}
arrange(activity, desc(steps))
```
Looking at the highest number of steps per interval, it seems like our subject works out early in the morning.


```{r prop, fig.width=6, fig.height=2, echo = "FALSE"}
# show proportion of movement to sedentary intervals
# Stacked barplot -- pretty hacky way of getting there

total <- nrow(activity_na_rm)

prop <- activity_na_rm %>%
  mutate(
    movement = steps > 0,
    movement2 = "active"
            ) %>%
  group_by(movement, movement2) %>%
  summarize(active = n(),
            prop = active / total
            )

# Calculate the cumulative sum of steps
prop <- plyr::ddply(prop, "movement2",
                   transform, label_pos=cumsum(active))

# Set label locations
prop$label_pos[1] <- 8900
prop$label_pos[2] <- 1450

# Plot proportions
ggplot(prop, aes(x=movement2, y=active, fill=movement)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Subject was active in 28% of intervals",
    x = NULL,
    y = "Intervals"
  ) +
# active label
  geom_text(data = tibble::as.tibble(split(prop, prop$movement)[[1]]), # use only one row of data
          aes(y=label_pos[1],
              x=movement2[1],
              label=stringr::str_c(as.character(scales::percent(prop[1])), "\nsedentary")
          ),
          hjust = 0) +
  # sedentary label
  geom_text(data = tibble::as.tibble(split(prop, prop$movement)[[2]]), # use only one row of data
            aes(y=label_pos[1],
                x=movement2[1],
                label=stringr::str_c(as.character(scales::percent(prop[1])), "\nactive")
            ),
            hjust = 0) +
  # nix ticks & hide legend
  theme(axis.text.y = element_blank(),
  axis.ticks.y = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  panel.background = element_blank(),
  plot.title = element_text(hjust = .09)) +
  theme(legend.position="none")
```

Here's a look at steps per day with missing values removed versus imputed.

```{r histo}
byday <- activity_na_rm %>%
  group_by(date) %>%
  summarise(steps = sum(steps)) 

ggplot(byday) +
geom_histogram(aes(steps), binwidth = 600) +
#  vertical x labels
theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
labs(title = "Subject achieved about 10,000 steps per day",
     subtitle = "Missing values removed",
  x = "number of steps/day",
  y = "number of occurences"
) +  # dashed median and mean lines
geom_vline(aes(
  xintercept = median(byday$steps, na.rm = T),
  color = "Median"),
    linetype = "dashed"
) +
geom_vline(aes(
  xintercept = mean(byday$steps, na.rm = T),
  color = "Mean"),
    linetype = "dashed"
) +
scale_y_continuous(breaks = c(1:10),
                   minor_breaks = NULL) +
 geom_text(data = tibble::tibble(steps = c(0, max(byday$steps)),
                date = c(1,2)),
             aes(y=9,
               x=13000,
               label=stringr::str_c("Median:\n", as.character(median(byday$steps, na.rm = T)))
            ),
           color = "#129496"
           ) +
geom_text(data = tibble::tibble(steps = c(0, max(byday$steps)),
                date = c(1,2)),
             aes(y=5,
               x=6000,
               label=stringr::str_c("Mean:\n", as.character(round(mean(byday$steps, na.rm = T), 2)))
            ),
           color = "#E7655C"
           ) +
  theme(legend.position="none")
```
  
```{r steps-day}
byday2 <- imputed %>%
  group_by(date) %>%
  summarise(steps = sum(steps)) 

ggplot(byday2) +
geom_histogram(aes(steps), binwidth = 600) +
#  vertical x labels
theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
labs(title = "Subject achieved about 10,000 steps per day",
     subtitle = "Missing values imputed",
  x = "number of steps/day",
  y = "number of occurences"
) +  # dashed median and mean lines
geom_vline(aes(
  xintercept = median(byday2$steps, na.rm = T),
  color = "Median"),
    linetype = "dashed"
) +
geom_vline(aes(
  xintercept = mean(byday2$steps, na.rm = T),
  color = "Mean"),
    linetype = "dashed"
) +
scale_y_continuous(breaks = c(1:10),
                   minor_breaks = NULL) +
 geom_text(data = tibble::tibble(steps = c(0, max(byday$steps)),
                date = c(1,2)),
             aes(y=6,
               x=13000,
               label=stringr::str_c("Median:\n", as.character(median(byday$steps, na.rm = T)))
            ),
           color = "#129496"
           ) +
geom_text(data = tibble::tibble(steps = c(0, max(byday2$steps)),
                date = c(1,2)),
             aes(y=6,
               x=7000,
               label=stringr::str_c("Mean:\n", as.character(round(mean(byday2$steps, na.rm = T), 2)))
            ),
           color = "#E7655C"
           ) +
  theme(legend.position="none")

```

The study's subject achieves the recommended [10,000 steps](https://www.ncbi.nlm.nih.gov/pubmed/11131268) on most days.

```{r long-time-series}
steps_seq <- activity
steps_seq <- tibble::rowid_to_column(steps_seq, "ID")

# steps_seq <- steps_seq %>%
#   mutate(hour = interval %/% 100,
#          minute = interval %% 60)

# I want to set up x axis tick labels with the number of days instead of intervals
# 288 5-minute intervals = 1 day
# Not sure how to get this to work

# intervals <- as.character(seq(from = 0, to = max(steps_seq$ID), by = 288))
# days <- as.character(seq(from = 0, to = 61, by = 1))
# day_labels <- vector("character", 62)
# day_labels <- for(i in seq_along(day_labels)) {
#   day_labels[i] <- stringr::str_c("'", intervals[i], "'", " = ", "'", days[i], "'")
# }

#  output should look like:
  # c("0" = "1", "288" = "2", "576" = "3", ...)

ggplot(data = steps_seq, aes(ID, steps)) + 
  geom_line() +
  labs(title = "Time series over 2-month study",
       subtitle = "Data were missing on 8 days",
  x = "number of steps/day",
  y = "number of occurences"
) +
  scale_x_discrete(breaks = c(seq(from = 0, to = max(steps_seq$ID), by = 288))
      # labels = day_labels  # still haven't figured out how to customize labels
                   ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5))
```

```{r weekend}
activity_na_rm %>%
  mutate(is.weekend = ifelse(day =="Saturday" | day =="Sunday", "weekend", "weekday"),
         time = lubridate::ymd_hm(paste0("2012-1-1", '"',as.character(hour),":", as.character(minute), '"'))) %>%

ggplot(aes(time,steps)) +
  geom_line() +
  facet_wrap(~is.weekend) +
  scale_x_datetime(date_breaks = "2 hours", date_labels = "%I:%M %p") +
  #  vertical x labels
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
  labs(title="More steps, earlier, on weekdays vs. weekends")
  

```


```{r histo-f-ckup, include=FALSE}
# excluding this chunk since it doesn't track days, according to the prompt

# zero-less

zeroless <- activity %>%
filter(steps != 0)

ggplot(zeroless) +
geom_histogram(aes(steps),
               bins = 100) +
scale_x_continuous(breaks = seq(0, max(activity$steps, na.rm = T), by = 25)) +
theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
labs(title = "Occurence of numbers of steps",
  subtitle = "(Zeros removed)",
  x = "number of steps",
  y = "number of intervals"
) +
geom_vline(aes(
  xintercept = median(zeroless$steps, na.rm = T),
  color = "Median"),
    linetype = "dashed"
) +
geom_vline(aes(
  xintercept = mean(zeroless$steps, na.rm = T),
  color = "Mean"),
    linetype = "dashed"
)

# zero-less square root scale
ggplot(zeroless) +
geom_histogram(aes(steps),
               bins = 100) +
scale_x_sqrt(
  breaks = c(
  seq(0, 25, by = 5),
  seq(50, 400, by = 25),
  seq(450, max(activity$steps, na.rm = T), by = 50)
  ),
  minor_breaks = c(
  seq(0, 45, by = 5),
  seq(50, 195, by = 5),
  seq(200, 480, by = 25),
  seq(450, max(activity$steps, na.rm = T), by = 25)
  )
) +
theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
labs(title = "Occurence of numbers of steps by 5-minute interval",
  subtitle = "(Zeros removed, square root scale)",
  x = "number of steps",
  y = "number of intervals"
) +
  geom_vline(aes(
  xintercept = median(zeroless$steps, na.rm = T),
  color = "Median"),
    linetype = "dashed"
) +
geom_vline(aes(
  xintercept = mean(zeroless$steps, na.rm = T),
  color = "Mean"),
    linetype = "dashed"
)
```